---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The code below reads the activity monitoring data:

```{r}
data <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?

The code below retrieves the total number of steps taken per day

```{r}
totalStepsPerDay <- tapply(data$steps, data$date, sum, na.rm=TRUE)
```

The results are as follows:
```{r echo=FALSE}
as.data.frame(totalStepsPerDay)
```

Below is a histogram for the total number of steps per day.

```{r}
hist(totalStepsPerDay, breaks=60)
```

Below is the code and results for computing the mean of the total number of steps per day.

```{r}
mean(totalStepsPerDay)
```

Below is the code and results for computing the median of the total number of steps per day.

```{r}
median(totalStepsPerDay)
```

## What is the average daily activity pattern?

The code and results below show the average number of steps taken per 5 minute interval across all days. 

```{r}
aveDailyActPattern <- tapply(data$steps, data$interval, mean, na.rm=TRUE)
aveDailyActPatternTable <- data.frame(as.numeric(aveDailyActPattern), as.numeric(names(aveDailyActPattern)))
plot(aveDailyActPatternTable[,2], aveDailyActPatternTable[,1], type="l", xlab="interval", ylab="average steps per interval")
```

The code and results below shows the 5 minute interval containing the maximum average number of steps.

```{r}
aveDailyActPatternTable[,2][aveDailyActPatternTable[,1] == max(aveDailyActPatternTable[,1], na.rm=TRUE)]
```

## Inputing missing values

The code and results below correspond to the number of NAs in the data.

```r
nrow(data[is.na(data$steps),])
```

To prevent bias in calculations, the code below replaces the missing values with the average steps for the specific 5 minute interval. The new data frame "dataFilled" without missing values is produced.

```{r}
dataFilled <- data
dataFilled$steps[is.na(dataFilled$steps)] <- aveDailyActPatternTable[,1][match(aveDailyActPatternTable[,2],dataFilled$interval)]
```

The code below computes the total steps per day based on the new data frame without missing values

```{r}
totalStepsPerDayFilled <- tapply(data$steps, data$date, sum, na.rm=TRUE)
as.data.frame(totalStepsPerDayFilled)
```

Below is a histogram for the total number of steps per day for the new data set without missing values.

```{r}
hist(totalStepsPerDayFilled, breaks=60)
```

Below is the code and results for computing the mean of the total number of steps per day for the new data set without missing values.

```{r}
mean(totalStepsPerDayFilled)
```

Below is the code and results for computing the median of the total number of steps per day for the new data set without missing values.

```{r}
median(totalStepsPerDayFilled)
```

Since the mean was used to fill up the missing values, we are not expecting any changes with the graph and summary statistics from the original data.

## Are there differences in activity patterns between weekdays and weekends?

The code below adds a factor indicating whether the date is a weekend or a weekday.

```{r results='hide'}
dataFilled[,4] <- ifelse(weekdays(as.Date(dataFilled[,2]))=="Saturday" | weekdays(as.Date(dataFilled[,2]))=="Sunday", "weekend", "weekday")
dataFilled[,4] <- as.factor(dataFilled[,4])
```

Below is the code and the generated time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r}
aveDailyActPatternWithDays <- tapply(dataFilled$steps, dataFilled[,c(3:4)], mean, na.rm=TRUE)
z <- cbind(as.numeric(rownames(aveDailyActPatternWithDays)), aveDailyActPatternWithDays[,1], aveDailyActPatternWithDays[,2])
par(mfrow=c(2,1))
par(mar=c(2,2,1,2))
plot(x = z[,1], y = z[,3], type="l", xlab = "", ylab = "", main="Weekend")
plot(x = z[,1], y = z[,2], type="l", xlab = "", ylab = "", main="Weekday")
```